Accelerating Metagenomic Identification of DNA Sequences Using Artificial Neural Networks \todo{Taxonomic Classification znaczy coś innego}

### 2025-04-15 article/main.tex
poprawki, rn
art, rn: poprawki, inna struktura

@@

    \title[Article Title]{Accelerating Taxonomic Classification of DNA Sequences Using Artificial Neural Networks}

@@

    \author*[1]{\fnm{Patryk Filip} \sur{Gryz}}\email{patryk.gryz.stud@pw.edu.pl}

    \affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{100190}, \state{State}, \country{Country}}}

    %%=============================================================%%
    %% ABSTRACT
    %%=============================================================%%

    \abstract{The rapid increase in the availability of DNA sequence data necessitates efficient and scalable methods for taxonomic classification. Traditional tools like BLAST, although accurate, often require significant computational resources and time, limiting their applicability to large datasets. Recent advances in artificial neural networks (ANNs) demonstrate their potential to address this issue by offering faster and more accurate solutions.

    This study introduces a novel ANN-based approach for the efficient taxonomic classification of DNA sequences. The proposed method reduces the complexity of large datasets by leveraging contrastive learning within an ANN to select representative sequences. This approach enhances computational efficiency by focusing classification efforts on these representatives rather than on the entire dataset. Comparative experiments were conducted on a benchmark dataset, evaluating the proposed ANN method alongside classical techniques such as k-mer embeddings and the Needleman-Wunsch algorithm.

@@

    \keywords{DNA sequencing, taxonomic classification, artificial neural networks, contrastive learning, sequence clustering}

@@

    \section{Background}
        The discovery of DNA marked the beginning of a new era in biology and medicine, enabling research into the molecular foundations of life and the precise diagnosis of many diseases\cite{Louie:2000}. In 1953, the structure of the DNA double helix was discovered by Watson and Crick\cite{Watson:1953}. In 1970, Francis Crick formulated the central dogma of molecular biology\cite{Crick:1970}, which states that genetic information flows from DNA to RNA and then to proteins, forming the foundation of modern molecular biology. In 1977, the complete genetic material of the DNA virus — bacteriophage $\phi{}$X174 was sequenced\cite{Sanger:1977:2}. This discovery highlighted the limitations of traditional analysis methods and demonstrated the need for computers to process DNA sequences\cite{Staden:1979}. In the 1980s, the European "EBML Data Library" was funded\cite{Higgins:1992}, and in 1982, the United States launched "GenBank"\cite{Bilofsky:1986}. In 1998, Jo Handelsman proposed the analysis of genetic material from many organisms simultaneously, without the need for cultivation, with his team studying organisms present in soil\cite{Handelsman:1998}. The use of information about genetic material found in the environment, combined with sequence databases, enables the identification of organisms present in a given sample. However, the process of analyzing large numbers of DNA sequences and comparing them to databases requires significant computational resources. With the development of next-generation sequencing methods\cite{Reinartz:2002}, sequencing costs have decreased significantly, and the number of DNA sequences being analyzed has increased\cite{Muir:2016}. Further accumulation of sequence data and increasing sequencing throughput lead to a growing demand for computational resources.

        \subsection{Motivation}
            The time required for taxonomic classification of large sequence datasets using traditional tools, such as BLAST, can span several hours, even for relatively small collections of sequences e.g. 4096 sequences. In many fields, artificial neural networks (ANN) have demonstrated the ability to outperform classical heuristic algorithms, offering faster and more accurate results. By leveraging the capabilities of ANN, this works aims to accelerate taxonomic classification process while maintaining at least the same level of quality as traditional methods.

        \subsection{Objectives}
            The primary objective of this study is to leverage an artificial neural network to support the selection of representative genetic sequences for taxonomic classification. Instead of classifying entire environmental genetic sequences sample, the proposed approach focuses on identifying a subset of sequences that best represent the sample. These selected representatives are then classified, reducing computational complexity while maintaining classification accuracy.

        % \subsection{Main Contributions}
        %     This paper introduces a novel artificial neural network model for calculating dissimilarities between genetic sequences, which can be used in clustering algorithm to select representative sequences for taxonomic classification. We provide all the necessary components for training the model, including dataset preparation, batch process, a loss function, and a training loop. The paper also includes an application that can perform taxonomic classification with specified parameters. Furthermore, we developed a library that allows users to compose the taxonomic classification pipelines.

        % \subsection{Structure of the Paper}
        %     The rest of the paper is organized as follows. Section 2 provides a description of related work, reviewing previous research on taxonomic classification and highlights a research gap. Section 3 defines the problem, presents the proposed approach, and provides detailed implementation details. Section 4 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis. Section 5 interprets the results, discusses the limtiations of proposed approach, and explores possible improvements for future work. Finally, Section 6 concludes the paper by summarizing the main findings and suggesting potential directions for future research.

        \subsection{Related Work}
            \subsection{Overview of Existing Approaches}
                Historically, one of the first algorithms that enabled taxonomic classification was the Needleman-Wunsch algorithm \cite{Needleman:1970}, developed in 1970, which allowed for the comparison of genetic sequences. However, the first solution that enabled taxonomic classification within a reasonable time frame was a tool created in 1983 by D. Lipman and W. Wilbur \cite{Wilbur:1983}. This method was based on dividing sequences into $k$-tuples, a generalization of $k$-mers, and comparing them. An advancement of this approach was the BLAST algorithm, introduced in 1990 \cite{Altschul:1990}, which also relied on $k$-mers and allowed for the efficient search of similar sequences in NCBI databases.

                The most significant progress in research, however, has occurred over the past 15 years. A notable example is the development of the MetaPhlAn tool in 2012 \cite{Segata:2012}, which uses marker genes to compare the species composition of metagenomic samples. Another marker gene-based approach is mOTUs2 \cite{Milanese:2019}, which identifies and analyzes markers unique to specific microbial strains. A different technique was employed in Centrifuge \cite{Kim:2016}, developed in 2016, which utilizes sequence indexing based on the Burrows-Wheeler transform \cite{Burrows:1994} and the Ferragina-Manzini index \cite{Ferragina:2000} for efficient sequence retrieval.

                A more classical approach is used in Kraken \cite{Wood:2014}, introduced in 2014, which combines $k$-mers with an indexed sequence database. An unconventional method for taxonomic classification was proposed in 2022 with BERTax \cite{Mock:2022}, which applies a transformer model \cite{Transformers} to DNA sequence analysis, treating DNA as a specialized language and allowing for classification without the need for reference databases. Another machine learning-based solution is CGRclust \cite{Alipour:2024}, which clusters DNA sequences using a two-dimensional chaos game representation and integrates unsupervised contrastive learning with convolutional neural networks.

            \subsection{Comparative Analysis}
                Algorithms based on $k$-mers are very fast, but they do not take into account all the information about the structure of the input genetic sequences, which reduces their quality. Solutions using marker genes are characterized by good quality, but they allow analysis only for subsets of organisms for which marker genes are known. Alignment-based tools should yield the best results, but they require significant computational resources.

            \subsection{Research Gaps}
                Although recent years have seen studies on taxonomic classification using machine learning methods, there is still a lack of research focusing on the application of artificial neural networks with contrastive learning for efficient sequence clustering. Such an approach could significantly accelerate the classification process using available tools.

@@

        \subsection{Proposed Approach}
            The proposed method involves reducing the dimensionality of input DNA sequences to a feature vector in the form of $\mathbb{R}^{64}$ using an artificial neural network. The neural network employs contrastive learning, which enables the learning of representations while preserving dissimilarity between sequences. The dissimilarity between sequence representations will be calculated using cosine dissimilarity, expressed by the formula in equation \eqref{Equation:CosineDissimilarity}.

            \begin{equation}
                dissimilarity_{cosine}(A, B) = 1 - similarity_{cosine}(A, B)
                \label{Equation:CosineDissimilarity}
            \end{equation}

            where:
            \begin{align*}
                similarity_{cosine}(A, B) =& \cos{\theta} = \frac{A \cdot B}{\|A\| \|B\|} = \frac{
                    \sum^{n}_{i = 1}A_i B_i
                }{
                    \sqrt{
                        \sum^{n}_{i = 1}A_i^2
                    }
                    \cdot
                    \sqrt{
                        \sum^{n}_{i = 1}B_i^2
                    }
                } \\
                A, B -& \text{compared vectors,} \\
                \theta -& \text{angle between $A$ and $B$,} \\
                A_j, B_j -& \text{$j$-th element of the vector $A$ and $B$, respectively.}
            \end{align*}

            Additionally, two classical approaches will be implemented in the work, which will be used for comparison with the neural network.

@@

@@

>>>>

\usepackage{todonotes}
\setuptodonotes{inline}

@@

\title[Metagenomic Identification using Neural Networks]{Accelerating Metagenomic Identification of DNA Sequences Using Artificial Neural Networks
\todo{Taxonomic Classification znaczy coś innego}
  }

@@

\author*[1]{\fnm{Patryk Filip} \sur{Gryz}}\email{patryk.gryz.stud@pw.edu.pl}
\author*[1]{\fnm{Robert} \sur{Nowak}}\email{robert.nowak@pw.edu.pl}

    \affil*[1]{\orgdiv{Institute of Computer Science}, \orgname{Warsaw University of Technology}, \orgaddress{\street{Nowowiejska 15/19}, \city{Warsaw}, \postcode{00-665}, \country{Poland}}}

    %%=============================================================%%
    %% ABSTRACT
    %% =============================================================%%

    \abstract{\todo{do przejrzenia, nie wszedzie poprawiony}The rapid increase in the availability of DNA sequence data necessitates efficient and scalable methods for metagenomic identification. Traditional tools like BLAST, although accurate, often require significant computational resources and time, limiting their applicability to large datasets. Recent advances in artificial neural networks (ANNs) demonstrate their potential to address this issue by offering faster and more accurate solutions.

    This study introduces a novel ANN-based approach for the efficient metagenomic identification of DNA sequences. The proposed method reduces the complexity of large datasets by leveraging contrastive learning within an ANN to select representative sequences. This approach enhances computational efficiency by focusing classification efforts on these representatives rather than on the entire dataset. Comparative experiments were conducted on a benchmark dataset, evaluating the proposed ANN method alongside classical techniques such as k-mer embeddings and the Needleman-Wunsch algorithm.

@@

    \keywords{DNA sequencing, metagenomic identification, artificial neural networks, contrastive learning, sequence clustering}

@@

    \section{Background}
    The use of information about genetic material found in the environment, combined with sequence databases, enables metagenomic identification -- the identification of organisms present in a given sample.
    With the development of next-generation sequencing methods\cite{Reinartz:2002}, sequencing costs have decreased significantly, and the number of DNA sequences being analyzed has increased\cite{Muir:2016}.
    The time required for metagenomic identification using traditional tools, such as BLAST, is unacceptable in many applications -- can span several hours, even for relatively small collections of analyzed sequences (for example 4096 sequences),
    due to huge size and explosive growth of sequencing databases \cite{10.1093/nar/gkr854}.
    This process require significant computational resources. It is necessary to create new methods to allow analyzing new samples of data in acceptable time using available resources.

    The primary objective of this study is to propose fast and high quality metagenomic identification method. We propose usage an artificial neural network to support the selection of representative genetic sequences.
    Instead of classifying entire environmental genetic sequences sample, the proposed approach focuses on identifying a subset of sequences that best represent the sample.
    These selected representatives are then used to identification, reducing computational complexity while maintaining accuracy.

    Historically, one of the first algorithms that enabled metagenomic identification was the Needleman-Wunsch algorithm \cite{Needleman:1970}, developed in 1970, which allowed for the comparison of genetic sequences.
    However, the first solution that enabled this task within a reasonable time frame was a tool created in 1983 by D. Lipman and W. Wilbur \cite{Wilbur:1983}.
    Their method was based on dividing sequences into $k$-tuples, a generalization of $k$-mers, and comparing them.
    An advancement of this approach was the BLAST algorithm, introduced in 1990 \cite{Altschul:1990}, which also relied on $k$-mers and allowed for the efficient search of similar sequences in sequence databases.

    The metagenomic ideitification became popular over the past 15 years, due to sequence database growth.
    A notable example of interesting tool is the development of the MetaPhlAn tool in 2012 \cite{Segata:2012}, which uses marker genes to compare the species composition of metagenomic samples.
    Another marker gene-based approach is mOTUs2 \cite{Milanese:2019}, which identifies and analyzes markers unique to specific microbial strains.

    A different technique was employed in Centrifuge \cite{Kim:2016}, developed in 2016,
    which utilizes sequence indexing based on the Burrows-Wheeler transform \cite{Burrows:1994} and the Ferragina-Manzini index \cite{Ferragina:2000} for efficient sequences retrieval.
    A more classical approach is used in Kraken \cite{Wood:2014}, introduced in 2014, which combines $k$-mers with an indexed sequence database.
    An unconventional method for metageonomic identification was proposed in 2022 with BERTax \cite{Mock:2022}, which applies a transformer model \cite{Transformers} to DNA sequence analysis,
    treating DNA as a specialized language and allowing for classification without the need for reference databases.
    Another machine learning-based solution is CGRclust \cite{Alipour:2024}, which clusters DNA sequences using a two-dimensional chaos game representation and integrates unsupervised contrastive learning with convolutional neural networks.

    Currently we can propose three types of methods developed for metagenomic identification: methods using alignment, methods based on $k$-mers and methods using merker genes.
    Alignment-based tools should yield the best results, but they require significant computational resources.
    Algorithms based on $k$-mers are very fast, but they do not take into account all the information about the structure of the input genetic sequences, which reduces their quality.
    Solutions using marker genes are characterized by good quality, but they allow analysis only for subsets of organisms for which marker genes are known.

    Although recent years have seen studies on metagenomic identification using machine learning methods, there is still
    a lack of research focusing on the application of artificial neural networks with contrastive learning for efficient sequence clustering.
    In many fields, artificial neural networks (ANN) have demonstrated the ability to outperform classical heuristic algorithms, offering faster and more accurate results.
    By leveraging the capabilities of ANN, this works aims to accelerate metagenomic identification process while maintaining at least the same level of quality as traditional methods.
    Such an approach could significantly accelerate the classification process using available tools.

    The rest of the paper is organized as follows. Section 2 defines the problem, presents the proposed approach, and provides detailed implementation details.
    Section 3 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis.
    Finally, section 4 concludes the paper by summarizing the main findings, discuss on the limtiations of proposed approach and suggesting potential directions for future research.

@@

    \todo{tutaj trzeba dac ogolny opis metody, moze rysunek, gdzie jest pokazane grupowanie, pytanie bazy danych itd}

    \todo{tutaj elementy systemu}

    \subsection{\todo{czy to nie jest wybor reprezentanta?} Proposed Approach}

    The proposed method involves reducing the dimensionality of input DNA sequences to a feature vector in the form of $\mathbb{R}^{64}$ using an artificial neural network.
    The neural network employs contrastive learning, which enables the learning of representations while preserving dissimilarity between sequences.
    The dissimilarity between sequence representations will be calculated using cosine dissimilarity, expressed by the formula in equation \eqref{Equation:CosineDissimilarity}.

            \begin{equation}
                dsim(A, B) = 1 - sim(A, B)
                \label{Equation:CosineDissimilarity}
            \end{equation}

            where:
            \begin{align*}
                sim(A, B) =& \cos{\theta} = \frac{A \cdot B}{\|A\| \|B\|} = \frac{
                    \sum^{n}_{i = 1}A_i B_i
                }{
                    \sqrt{
                        \sum^{n}_{i = 1}A_i^2
                    }
                    \cdot
                    \sqrt{
                        \sum^{n}_{i = 1}B_i^2
                    }
                } \\
                A, B -& \text{compared vectors,} \\
                \theta -& \text{angle between $A$ and $B$,} \\
                A_j, B_j -& \text{$j$-th element of the vector $A$ and $B$, respectively.}
            \end{align*}

@@

        \subsection{Classical approach}

                \todo{tutaj metody klasyczne, porownawcze}

            Two classical approaches were implemented and used for comparison with the proposed neural network.

@@

\subsubsection*{Author contributions}

P.G and R.N. identified the problem, P.G. designed the approach, downloaded the data, implemented the software, performed numerical experiments, P.G. and R.N. prepared the draft. All authors have read and agreed to the published version of the manuscript.

\subsubsection*{Funding}

A statutory Research Grant from the Institute of Computer Science, Warsaw University of Technology, supports this work.
